---
title: "week4_homework"
output: html_document
---

#read in the csv gender inequality and the geojson world map 

```{r}
library(sf)
library(tidyverse)
library(here)
library(dplyr)
library(janitor)
library(readr)
library(tidyr)
library(countrycode)
library(tmap)

composite_indices <- read.csv('composite_indices.csv')

world_map <- st_read('World_Countries.geojson')

```

#filter the columns so only Gender Inequality Index for 2010 and 2019 are left for each country

```{r}
gender_inequality <- composite_indices %>% 
  group_by(country) %>% 
  select(iso3, country, gii_2010, gii_2019)

gender_inequality
```

#Now, create an extra column that looks at the gii difference between 2019 and 2010

```{r}
gender_inequality_diff <- gender_inequality %>% 
  mutate(gii_diff = (gii_2019 - gii_2010))

gender_inequality_diff
```

#Now, we have to join the World Map data with the new Gender Inequality Diff table - however, we have to make the column iso3 codes the same. For that, we'll use the countrycode package

```{r}
#transform the country code in the gender inequality file to match that of the world map file
#then remove all the regions/non-country rows

gender_inequality_diff$iso2 <- countrycode(gender_inequality_diff$iso3, origin = "iso3c", destination = "iso2c") 

gender_inequality_diff <- gender_inequality_diff %>% 
  filter(iso2 != "NA")

gender_inequality_diff


```

#Join the world map file to the gender inequality updated file

```{r}
gender_inequality_diff_map <- world_map %>% 
  inner_join(.,
             gender_inequality_diff,
             by = c("ISO" = "iso2"))

gender_inequality_diff_map
```

#Plot the map! We'll use the "view" mode to make it interactive

```{r}
tmap_mode("view")

tm_shape(gender_inequality_diff_map) +
  tm_polygons("gii_diff",
              palette = "-Spectral",
              title = "GII Difference, 2010-2019",
              id = "COUNTRY",
              alpha = 0.5,
              midpoint = NA) 

```


#If we ant to use plot mode, we can use the following code - added in code to make the legend smaller

```{r}
tmap_mode("plot")

tm_shape(gender_inequality_diff_map) +
  tm_polygons("gii_diff",
              palette = "-Spectral",
              title = "GII Difference, 2010-2019",
              id = "COUNTRY",
              alpha = 0.5,
              midpoint = NA) +
  tm_legend(
    legend.text.size = 0.6, 
    legend.title.size = 0.8,
    legend.position = c("left", "bottom")
  )
```


